---
title: 'Worksheet 6: Topic Modeling'
author: 'Antonio Worley'
date: '04/02/2025'
---

_This is the sixth in a series of worksheets for History 8500 at Clemson University. The goal of these worksheets is simple: practice, practice, practice. The worksheet introduces concepts and techniques and includes prompts for you to practice in this interactive document. When you are finished, you should change the author name (above), render your document to a pdf, and upload it to canvas. Don't forget to commit your changes as you go and push to github when you finish the worksheet._

Text analysis is an umbrella for a number of different methodologies. Generally speaking, it involves taking a set (or corpus) of textual sources, turning them into data that a computer can understand, and then running calculations and algorithms using that data. Typically, at its most basic level, that involves the counting of words.

Topic modeling (TM) is one type of text analysis that is particularly useful for historians. 

TM takes collections or corpuses of documents and returns groups of "topics" from those documents. It is a form of unsupervised classification that finds groups of items that are probabilistically likely to co-occur. 

Latent Dirichlet allocation (LDA) is the most popular algorithm or method for topic modeling, although there are others. It assumes that each document has a mixture of topics and that each topic is a mixture of words. That means that topics overlap each other in terms of content rather than being confined to distinct and singular groups. 

To prepare a corpus for topic modeling, we'll do many of the same types of operations that we used last week to prepare a corpus for analysis. First we'll pre-process the data and then we'll create a document term matrix from our corpus using the `tm` (text mining) package. 

```{r}
library(tidytext)
library(tidyverse)
library(readtext)
library(tm)
library(topicmodels)

```

```{r}
download.file("https://github.com/regan008/8510-TextAnalysisData/blob/main/TheAmericanCity.zip?raw=true", "AmCity.zip")
unzip("AmCity.zip")
```

```{r}
# Metadata that includes info about each issue.
metadata <- read.csv("https://raw.githubusercontent.com/regan008/8510-TextAnalysisData/main/AmCityMetadata.csv")

meta <- as.data.frame(metadata)
#meta$Filename <- paste("MB_", meta$Filename, sep="")

ac_texts <- readtext(paste("/home/antonioworley70/TheAmericanCity/*.txt"))
ac_whole <- full_join(meta, ac_texts, by = c("filename" = "doc_id")) %>% as_tibble() 
tidy_ac <- ac_whole %>%
  unnest_tokens(word, text) %>% 
  filter(str_detect(word, "[a-z']$")) %>% 
  anti_join(stop_words)

tidy_ac <- tidy_ac %>% filter(!grepl('[0-9]', word))

```
The above code borrows from what we did last week. It pulls in the texts from the _The American City_ corpus, joins them together into a single data frame, and then turns then uses `unnest_tokens()` to tokenize the text and, finally, removes stop words. 

For topic modeling, we need a Document Term Matrix, or a DTM. Topic Modeling has the documents running down one side and the terms across the top. `Tidytext` provides a function for converting to and from DTMs. First, we need to create a document that has the doc_id, the word and the count of the number of times that word occurs. We can do that using `count()`.

```{r}
tidy_ac_words <- tidy_ac %>% count(filename, word)
```

Now we can use `cast_dtm()` to turn `tidy_mb_words` into a dtm. 

```{r}
ac.dtm <- tidy_ac_words %>% 
  count(filename, word) %>% 
  cast_dtm(filename, word, n)
```

If you run `class(mb.dtm)` in your console you will notice that it now has a class of "DocumentTermMatrix". 

Now that we have a dtm, we can create a topic model. For this, we'll use the topic models package and the `LDA()` function. Take a minute and read the documentation for `LDA()`.

There are two important options when running `LDA()`. The first is k which is the number of topics you want the model to generate. What number topics you generate is a decision that often takes some experimentation and depends on the size of your corpus. The American City corpus isn't that bigbut still has over 209k words. In this instance, because the corpus is so small we're going to start with a small number of topics. Going above 5 causes errors with this particular corpus. Later, when you work with a different corpus you should experiment with changing the number of topics from 10 to 20 to 30 to 50 to see how it changes your model. 

The second important option when running `LDA()` is the seed option. You don't worry too much about what setting the seed does, but put simply - it ensures the output of the model is predictable and reproducible. Using the seed ensures that if you come back to your code later or someone else tries to run it, the model will return exactly the same results. 

Lets now train our model. This will take a few minutes: 
```{r}
ac.lda <- LDA(ac.dtm, k = 5, control = list(seed = 12345))
```

Now we have a LDA topic model that has 5 topics. There are two ways to look at this model: word-topic probabilities and document-topic probabilities. 

Lets start with **word-topic probabilities.**

Every topic is made up of words that are most associated with that topic. Together these words typically form some sort of theme. To understand what this looks like the easiest thing to do is create a bar chart of the top terms in a topic. 

```{r}
ac.topics <- tidy(ac.lda, matrix = "beta")
head(ac.topics)
```
What we have here is a list of topics and the weight of each term in that topic. Essential we have turned this into a one-topic-per-term-per-row format. So, for example, the term 10th has a weight of 5.135047e-05 in topic 1 but 7.269700e-05 in topic 2. Now that doesn't mean a lot to us at this moment and this format is impossible to grasp in its current size and iteration, but we can use tidyverse functions to pair this down and determine the 10 terms that are most common within each topic. 
```{r}
ac.top.terms <- ac.topics %>%
  arrange(desc(beta)) %>% 
  group_by(topic) %>% slice(1:5)
  x11()
ac.top.terms %>%
  mutate(term = reorder_within(term, beta, topic)) %>%
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered()
```
(@) Can you adjust the code above to show the top 10 words from just one topic?
```{r}
ac.top.terms <- ac.topics %>%
  arrange(desc(beta)) %>% 
  group_by(topic) %>% slice(1:10)
  x11()
ac.top.terms %>%
  mutate(term = reorder_within(term, beta, topic)) %>%
  filter(topic == "1") %>%
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered()

```

Another useful way to look at the words in each topic is by visualizing them as a wordcloud.
```{r warning=FALSE}
library(wordcloud)
x11()
topic1 <- ac.topics %>% filter(topic == 2)
wordcloud(topic1$term, topic1$beta, max.words = 100, random.order = FALSE,
    rot.per = 0.3, colors = brewer.pal(6, "Dark2"))
```


Now we can see what words are most common in each topic. But the document-topic probabilities are also useful for understanding what topics are prevalent in what documents. Just as each topic is made up of a mixture of words, the LDA algorithm also assumes that each topic is made up of a mixture of topics. 

```{r}
ac.documents <- tidy(ac.lda, matrix = "gamma")
head(ac.documents)
```
For each document, the model gives us an estimated proportion of what words in the document are from a topic. So for the April 1915 issue it estimates that about 23% of the words are from topic 1. The gamma number represents the posterior topic distribution for each document. 

This is easier to see if we filter to see the breakdown for just one document. 
```{r}
ac.documents %>%  filter(document == "1916_May.txt") %>% arrange(desc(gamma))
```

This gamma value is really useful and we can use it to see which topics appear in which documents the most. This is frequently referred to as looking at topics over time. 

We can do that using the ac.documents dataframe that we just created but it needs to be joined with the metadata. Again, this is why it is important to have a filename within the metadata spreadsheet. To join these two together we can do a full_join because we want to keep all of the columns.
```{r}
topics.by.year <- full_join(ac.documents, metadata, by = join_by(document == filename))
```

Now what we have is a document that includes the proportion of each topic in each document. Because this is a dataset about a periodical, we have values in our metadata that will make it easy to plot the distrubtion of a topic over time -- in this case for each edition of the journal.
```{r}
topics.by.year$issue_date <- paste(topics.by.year$month, " ", topics.by.year$year, sep = "")
x11()
ggplot(data=topics.by.year, aes(x=issue_date, y=gamma)) + geom_bar(stat="identity") + facet_wrap(~ topic, scales = "free") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Using this we can tell that topic 5, which from earlier had the words improve, grow, ties, contracts, and gasoline as the top five words, is most prominent in January 1915. 

(@) Use the rest of this worksheet to experiment with topic modeling. I've added the code to download a much larger dataset - the issues of Mind and Body. This corpus has 413 documents ranging from the 1890s to 1936. You'll want to start with at least 25 topics. 
```{r}
#| eval: false
download.file("https://github.com/regan008/8510-TextAnalysisData/blob/main/MindAndBody.zip?raw=true", "C:/Users/anton/Desktop/MB.zip")
unzip("MB.zip")

```

```{r}
# Metadata that includes info about each issue.
mb.metadata <- read.csv("https://raw.githubusercontent.com/regan008/8510-TextAnalysisData/main/mb-metadata.csv")
mb<-readtext(paste("C:/Users/anton/Desktop/txt/*.txt"))

mb<- mb %>%
  mutate(doc_id = gsub("MB_", "", doc_id))

full.mb<- mb%>%
  full_join(mb.metadata, by = join_by(doc_id == Filename))

mb.tokens<- full.mb %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words) 

mb.tokens<- mb.tokens %>%
  filter(!grepl("[0-9]", word))
 

mb.dtm <- mb.tokens %>% 
  count(doc_id, word) %>% 
  cast_dtm(doc_id, word, n)


mb.lda<- LDA(mb.dtm, k = 25, control = list(seed = 12345), method = "Gibbs", alpha = .05)

mb.topics <- tidy(mb.lda, matrix = "beta")
mb.top.terms <- mb.topics %>%
  arrange(desc(beta)) %>% 
  group_by(topic) %>% slice(1:5)
  View(mb.top.terms)
  #topic 1 I don't really know but maybe fencing? topic 2 seems like nothing, topic 3 something to do with being outside, topic 4 food, topic 5 no idea, topic 6 I think it's something(like not random words) but no idea, topic 7 clearly measurements, topic 8 health but to do with heart rate and stuff, topic 10 is gym/playground, topic 11 health physical education, topic 14 is all verbs so probably instructions, topic 18 has cities so that's interesting, topic 25 has digitized so that's weird. There is a clear physical, health, and workout theme here though, which of course makes sense considering it's a physical health magazine. 

```

(@) What happens if you create a custom stopword list? How does this change the model?
```{r}
mystops<- c("forw", "backw", "cts", "bd", "st", "ave", "pos", "yi")
mb.tokens<- mb.tokens %>%
  filter(!word %in% mystops)

mb.dtm <- mb.tokens %>% 
  count(doc_id, word) %>% 
  cast_dtm(doc_id, word, n)


mb.lda<- LDA(mb.dtm, k = 25, control = list(seed = 12345), method = "Gibbs", alpha = .05)

mb.topics <- tidy(mb.lda, matrix = "beta")
mb.top.terms <- mb.topics %>%
  arrange(desc(beta)) %>% 
  group_by(topic) %>% slice(1:5)
  View(mb.top.terms)  
#this drastically changed the topics, topic 1 is about robe climbing now, 2 is excercises, and 3 is skeletal-muscular, it has completely shifted the order of topics but most of the composition is the same, a few new words like topic 9 has santa stage king and sing which seems like it's about a christmas parade. 
```


(@) Can you create a topic model for just the documents in the 1920s? How does that change the model? 
```{r}
mb.tokens<- mb.tokens %>%
  filter(Year == 1920)

mb.dtm <- mb.tokens %>% 
  count(doc_id, word) %>% 
  cast_dtm(doc_id, word, n)


mb.lda<- LDA(mb.dtm, k = 25, control = list(seed = 12345), method = "Gibbs", alpha = .05)

mb.topics <- tidy(mb.lda, matrix = "beta")
mb.top.terms <- mb.topics %>%
  arrange(desc(beta)) %>% 
  group_by(topic) %>% slice(1:5)
  View(mb.top.terms) 


# The topics are clearly different but some still have to do with physical education. The word racial appears in topic 25, I'm not entirely certain why that would be especially prevalent in 1920. There is also  "girls" are in multiple topics and this could be because of the ratification of the 19th amendment. Perhaps this is because women and girls were simply at the front, or at least more at the front, of society's mind so the journal began covering girls. 

```



(@) Finally, when you are all done. Write up your findings. What research question did you focus on and what did you learn? 
```{r}

txt<- readtext("C:/Users/anton/Desktop/herald1874.txt")
#this is a single page of the New York Herald, I realize that a single page topic model may not work however this page consists solely of quotes and political opinions from every senator and many representatives from 1874 about President Grant's possible 3rd term. I thought it would be fascinating to see if any patterns emerge, it may point to underlying political attitudes or suprising divides or overlaps between democrats and republicans.

herald.tokens<- txt %>%
  unnest_tokens(sentence, text, token = "sentences") 
#decided to group by sentences then group by number of sentences because the page is a wall of text
 #doesn't make sense to do stop words when im doing sentences at this stage
  
my_stops<- c("1", "ami", "street", "tho", "ol", "1", "2", "3", "4", "5", "6", "7", "8", "9", "lor", "o\\'clock", "bo", "aud", "day", "10", "st", "west", "john", "oi", "de", "ac", "ho", "12","tne", "11", "20", "tue", "lie", "tlie", "lo", "iu", "dr", "con", "ou", "01", "irom", "ot", "0", "la", "tor", "100", "ii", "tbe", "william", "lu","111", "ft", "city", "york", "twenty", "li","ten","time","late","house", "sale", "herald", "address", "east", "morning","30", "200","13", "fourth", "si", "16", "called", "18", "thirty", "ing", "avenue", "evening", "ana", "held", "17", "15", "1874", "00", "50", "14", "1873", "night", "yesterday","office","broadway","board","sixth","hall", "ol," "tbe", "oi")
my_stops_clean <- my_stops %>%
  tolower() %>%        
  trimws() %>%        
  gsub("[^a-z]", "", .)
  #some things in my stops weren't getting applied and AI suggested using this because unnest tokens does certain things to text like lowercasing and stuff so my_stops was not in the correct format to apply the stops
 herald.tokens<- herald.tokens %>%
  mutate(group = (row_number() - 1) %/% 6) %>%
  #only way I could figure to do this, this creates groups by every 3 rows, I'm gonna try various groupings
  #6 sentences seems to give me the most meaningful topics, in the original page, quotes are broken up by 3-7 lines so I think this works
  group_by(group) %>%
  summarise(text = paste(sentence, collapse = " ")) 
  #paste the sentence rows into one


tokens<- herald.tokens %>%
  unnest_tokens(word,text) %>%
    anti_join(stop_words) %>%
     filter(!word %in% my_stops_clean)



herald.dtm <- tokens %>% 
  count(group, word) %>% 
  cast_dtm(group, word, n)


herald.lda<- LDA(herald.dtm, k = 25, control = list(seed = 12345), method = "Gibbs", alpha = .05)
#need to play with topics as much as possible
#ran 3-10 topics, most were very general and not that informative
#4 was the cleanest and 25 had the most interesting things
herald.topics <- tidy(herald.lda, matrix = "beta")

herald.top.terms <- herald.topics %>%
  arrange(desc(beta)) %>% 
  group_by(topic) %>% slice(1:5)
  View(herald.top.terms)
  x11()
ggplot(data=herald.top.terms, aes(x=term, y=beta)) + geom_bar(stat="identity") + facet_wrap(~ topic, scales = "free") + 
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


#Oddly enough 4 topics gave me the most sensible collection of words. Topic 1 had republican, grant, support, opposed, nominated and 3 had democrat, term, republican, people, and opinion. Now topic 1 is extremely fascinating because the model is linking republican with support AND opposed to the prospect of Grant's third term. In 1874 there was certainly a disharmony in the republican party and this seems to be somewhat reflecting that. Another odd turn of events, I did 25 topics and found extremely relevant groupings. Though it was filled with ocr errors, many of the errors were actually readable words, like urant for grant. You have words like support with specific states, you have states grouped together under topic, you have a topic with reason, grant's, south, people, and also multiple topics with Kentucky. Grant had several ties to Kentucky so that's either relevant or an odd coincidence. Topic 3 has democrat, republican, would, and opposed which is interesting. The topics seem to be indicating that certain political issues could cut across party lines. The way the model is grouping creates avenues for inquiry that I wouldn't have thought to pursue. Now is this a usable form of digital argument? Probably not, however it does raise interesting questions. 

```



**This was the final worksheet for this class. You did it - you learned to code in R! Congrats!**

